Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2020-09-02 journal paper/content/sec-intro.tex
Revising the intro ...

Learning from interactions with users is ubiquitous in modern customer-facing platforms, from product recommendations to web search to content selection to fine-tuning user interfaces. Many platforms purposefully implement \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information.
Online platforms routinely deploy A/B tests, and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a standard and well-studied framework for exploration and making decisions under uncertainty
\citep{Gittins-book11,Bubeck-survey12,slivkins-MABbook,LS19bandit-book}.

@@

In this paper, we initiate a study of the interplay between \exploration and \competition. Platforms that engage in exploration typically need to compete against one another; most importantly, they compete for users. This creates a tension:
%between \exploration and \competition.
while exploration may be essential for improving the service tomorrow, it may degrade quality and make users leave \emph{today}, in which case there will be less users to learn from. This may further degrade the platform's performance relative to competitors who keep learning and improving from \emph{their} users, and so forth. Taken to the extreme, such dynamics may create a ``death spiral" effect when the vast majority of customers eventually switch to competitors. Users therefore serve three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents who choose among the competing systems.

The main high-level question that we focus on in this paper is:
\begin{align}\label{eq:main-Q}
\textbf{How does competition incentivize the adoption of better exploration algorithms?}
\end{align}

\noindent This translates into a number of more concrete questions. While it is commonly assumed that better technology always helps, is this so under competition? Does increased competition lead to higher consumer welfare? How significant are ``data feedback loops" --- when more data leads to more users, which leads to even more data, etc. --- and how they relate to the anti-trust considerations? Finding formalizations that admit meaningful answers is a major part of the overall challenge.

To answer these questions, we study complex interactions between platforms' learning dynamics and users' self-interested behavior. The choice of a particular technology (exploration algorithm) is no longer an abstract, static choice with a predetermined outcome for the platform. Instead, we model the algorithms explicitly, and investigate how they play out in competition over an extended period of time.

@@

\subsection{Our model: competing bandits}
\label{sec:intro-model}

%We investigate these questions with
\xhdr{Competition game.} We consider a stylized duopoly model where two firms commit to exploration strategies and compete for a stream of consumers. We define a game in which two firms (\emph{principals}) simultaneously engage in exploration and compete for users (\emph{agents}). These two processes are interlinked, as exploration decisions are experienced by users and informed by their feedback. We need to specify several conceptual pieces: how the principals and agents interact, what is the machine learning problem faced by each principal, and what is the information structure. Each piece can get rather complicated in isolation, let alone jointly, so we strive for simplicity. Thus, the key features of our model are as follows:

\begin{itemize}

\item A new agent arrives in each round $t=1,2, \ldots$, and chooses among the two principals. The principal chooses an action (\eg a list of web search results to show to the agent), the user experiences this action, and reports a reward. All agents have the same ``decision rule" for choosing among the principals given the available information.

\item Each principal faces a basic and well-studied version of the multi-armed bandit problem: for each arriving agent, it chooses from a fixed set of actions  (a.k.a. \emph{arms}) and receives a reward drawn independently from a fixed distribution specific to this action. The reward distributions are initially unknown (but can be estimated over time from the data).

\item Principals simultaneously announce their bandit algorithms before round $1$, and cannot change them afterwards.  Each principal's objective is to maximize its market share (the fraction of users choosing this principal). Each principal only observes agents that chose this principal.
%, but may have access to each platforms' reputation score (more on this in    Section~\ref{sec:intro-discussion}).
\end{itemize}

\noindent We investigate several model variants within this framework, where we vary agents' decision rule and/or give a first-mover advantage to one of the principals. In all variants, agents have little or no information about other agents' choices and rewards.

@@

>>>>

Learning from interactions with users is ubiquitous in modern customer-facing platforms, from product recommendations to web search to content selection to fine-tuning user interfaces. Many platforms purposefully implement \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information.
Online platforms routinely deploy A/B tests, and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a standard and well-studied framework for exploration and making decisions under uncertainty.
%\citep{Gittins-book11,Bubeck-survey12,slivkins-MABbook,LS19bandit-book}.

@@

In this paper, we study the interplay between \exploration and \competition. Platforms that engage in exploration typically need to compete against one another; most importantly, they compete for users. This creates a tension:
%between \exploration and \competition.
while exploration may be essential for improving the service tomorrow, it may degrade quality and make users leave \emph{today}, in which case there will be less users to learn from. This may further degrade the platform's performance relative to competitors who keep learning and improving from \emph{their} users, and so forth. Taken to the extreme, such dynamics may create a ``death spiral" effect when the vast majority of customers eventually switch to competitors. Users therefore serve three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents who choose among the competing systems.

The main high-level question is: 
%that we focus on in this paper is:
%\begin{align}\label{eq:main-Q}
\textbf{How does competition incentivize the adoption of better exploration algorithms?}
%\end{align}
This translates into a number of more concrete questions. While it is commonly assumed that better technology always helps, is this so under competition? Does increased competition lead to higher consumer welfare? How significant are ``data feedback loops" --- when more data leads to more users, which leads to even more data, etc. --- and how they relate to the anti-trust considerations? 
%Finding formalizations that admit meaningful answers is a major part of the overall challenge.
To answer these questions, we study complex interactions between platforms' learning dynamics and users' self-interested behavior. The choice of a particular technology (exploration algorithm) is no longer an abstract, static choice with a predetermined outcome for the platform. Instead, we model the algorithms explicitly, and investigate how they play out in competition over an extended period of time. We offer a mix of theoretical results and numerical simulations.

@@

%\subsection{Our model: competing bandits}
%\label{sec:intro-model}

%We investigate these questions with
\xhdr{Our model: Competition game.} We consider a stylized duopoly model where two firms commit to exploration strategies and compete for a stream of consumers. We define a game in which two firms (\emph{principals}) simultaneously engage in exploration and compete for users (\emph{agents}). These two processes are interlinked, as exploration decisions are experienced by users and informed by their feedback. We need to specify several conceptual pieces: how the principals and agents interact, what is the machine learning problem faced by each principal, and what is the information structure. Each piece can get rather complicated in isolation, let alone jointly, so we strive for simplicity. Thus, the key features of our model are as follows:

%\begin{itemize}

\textbf{(i)} A new agent arrives in each round $t=1,2, \ldots$, and chooses among the two principals. The principal chooses an action (\eg a list of web search results to show to the agent), the user experiences this action, and reports a reward. All agents have the same ``decision rule" for choosing among the principals given the available information.

\textbf{(ii)} Each principal faces a basic and well-studied version of the multi-armed bandit problem: for each arriving agent, it chooses from a fixed set of actions  (a.k.a. \emph{arms}) and receives a reward drawn independently from a fixed distribution specific to this action. The reward distributions are initially unknown.
% (but can be estimated over time from the data).

\textbf{(iii)} Principals simultaneously announce their bandit algorithms before round $1$, and cannot change them afterwards.  Each principal's objective is to maximize its market share (the fraction of users choosing this principal). Each principal only observes agents that chose this principal.
%, but may have access to each platforms' reputation score (more on this in    Section~\ref{sec:intro-discussion}).
%\end{itemize}

We investigate several model variants within this framework, where we vary agents' decision rule and/or give a first-mover advantage to one of the principals. In all variants, agents have little or no information about other agents' choices and rewards.

@@

\ascomment{revising ...}

\xhdr{Technology: multi-armed bandit algorithms.}
To compare between bandit algorithms, we build on prevalent intuition in the literature. We focus on standard notions of regret, and distinguish between three classes of bandit algorithms: ones that never explicitly explore (\emph{greedy algorithms}), ones that explore without looking at the data (\emph{exploration-separating algorithms}), and ones where exploration gradually zooms in on the best arm (\emph{adaptive-exploration algorithms}). In the absence of competition, these classes are fairly well-understood: greedy algorithms are terrible for a wide variety of problem instances, exploration-separated algorithms learn at a reasonable but mediocre rate across all problem instances, and adaptive-exploration algorithms are optimal in the worst case, and exponentially improve for ``easy" problem instances. Generally,  ``better" algorithms are better in the long run, but could be worse initially.
